A unified pseudo-$C_\ell$ framework

### 2018-08-22 doc/Paper/main.tex

      At this stage, the pseudo-$C_\ell$ method proceeds as usual, applying Eq. \ref{eq:pcldef} to the purified field. The purification process however requires a slight modification of the analytical form of the mode-coupling matrix. Specifically, if either $E$ or $B$ modes of a spin-2 field have been purified Eqs. \ref{eq:m02} and \ref{eq:m22} must be modified by exchanging one factor of
      \begin{equation}\nonumber
        \wtj{\ell}{\ell'}{\ell''}{2}{-2}{0}
      \end{equation}
      for
      \begin{align}\nonumber
        &\wtj{\ell}{\ell'}{\ell''}{2}{-2}{0}+2\frac{\beta_{\ell,2}}{\beta_{\ell,1}\beta_{\ell'',1}}\wtj{\ell}{\ell'}{\ell''}{1}{-2}{1}\\
        &\,\,\,+\frac{\beta_{\ell,2}}{\beta_{\ell'',2}}\wtj{\ell}{\ell'}{\ell''}{0}{-2}{2}.
      \end{align}
      per purified field. \todo{check this}.

@@

      The full covariance matrix is very close to diagonal, as shown
      in the right panel of Fig. \ref{fig:lss_val_cls}. Under the
      assumption that the estimator is unbiased
      ($\langle{\bf s}\rangle={\bf t}$), the values of $\chi^2_i$
      should follow a $\chi^2$ distribution with a $N_{\rm dof}$
      degrees of freedom. Figure \ref{fig:lss_val_chi2} shows the
      distribution of $\chi^2_i$ across simulations approximating the
      covariance matrix as diagonal (blue histograms),
      compared with the predicted $\chi^2$ distribution for
      $N_{\rm dof}=681$ (marked by the vertical dashed lines) for the
      six different power spectrum combinations. Both distributions
      are found to be compatible. Moreover, the dot-dashed lines show
      the values of $\chi^2$ found for the mean residual with respect
      to the input power spectra (middle panel of
      Fig. \ref{fig:lss_val_cls}). The probability to exceed (PTE)
      associated with these $\chi^2$ values are higher than $10\%$ in
      all cases, and therefore our validation suite does not show any
      statistically significant bias in the estimator. This bias, if
      any, must be at least $\sqrt{1000}\sim30$ times smaller than the
      statistical uncertainties associated with our validation suite,
      which are representative of the statistical power of
      next-generation galaxy surveys.

@@

    \subsection{Curved-sky CMB \texorpdfstring{$B$}{B}-mode validation}\label{ssec:validation.cmb_full}
      \begin{figure*}
        \centering
        \includegraphics[width=0.99\columnwidth]{val_covar_cmb_sph_nocont}
        \includegraphics[width=0.99\columnwidth]{val_covar_cmb_sph}
        \caption{{\sl Left:} covariance matrix for power spectra computed in the absence of foreground contaminants computed from the 1000 simulations of our curved-sky CMB validation suite. The incomplete sky coverage causes a noticeable anti-correlation of neighboring bandpowers at low $\ell$ for the three power spectra ($EE$, $EB$ and $BB$), but beyond that, the covariance matrix is close to diagonal. {\sl Right:} the same as the left panel but this time for simulations containing foreground contaminants. The combination of $B$-mode purification and contaminant deprojection produces noticeable correlations at low $\ell$, especially in the $BB$ spectrum, that must be taken into account.} \label{fig:cmb_val_cov}
      \end{figure*}
      \begin{figure*}
        \centering
        \includegraphics[width=0.80\textwidth]{val_chi2_cmb_sph}
        \caption{Distribution of $\chi^2$ values found for the 1000 simulations in our curved-sky CMB validation suite (blue histograms). Results are shown for the three polarized spectra ($EE$, $EB$ and $BB$), labeled in the top right corner of each panel. The simulated distributions agree well with the expected $\chi^2$ distribution (solid line) assuming Gaussian statistics for $N_{\rm dof}=43$ degrees of freedom (denoted by the dashed vertical lines). The dot-dashed vertical lines mark the $\chi^2$ value found for the mean residual with respect to the input power spectrum (see lower half of the left panel in Fig. \ref{fig:cmb_val_cls}). The associated PTE values are $\gtrsim0.4$, and thus no statistically significant bias is found by our validation suite.} \label{fig:cmb_val_chi2}
      \end{figure*}
      We have carried out a similar set of tests on an ensemble of
      1000 simulations from the CMB validation suite described in
      Section \ref{sssec:validation.suite.cmb}. In our fiducial run we
      both deprojected the foreground contaminants and subsequently
      used $B$-mode purification to decouple $E$ and $B$-modes at the
      map level, since this is the main feature of the code we want to
      validate. The left panel of Figure \ref{fig:cmb_val_cls} shows
      the mean of the $EE$, $EB$ and $BB$ power spectra computed for
      all simulations (solid lines) compared with the expectation from
      the input spectra for $EE$ and $BB$ (dashed lines, mostly lying
      under the solid ones). The average residuals normalized over the
      error on the mean, shown in the lower panel of the same Figure,
      fluctuate around zero with an amplitude of up to $3\sigma$, as
      would be expected in the absence of a bias in the
      estimator.

      The above figure was created using E/B mode
      purification. Disabling this option has no impact on the bias of
      the estimates, but increases its variance as we show in the
      Figure \ref{fig:cmb_val_sigma}. We plot diagonal error bars with (red) and without (blue) $B$-mode purification for the $BB$ (solid), $EB$ (dashed) and $EE$ (dot-dashed) power spectra. Although there is no practical improvement in the error bars for the $EE$ spectrum (there is even a slight increase due to the loss of the ambiguous $E$ modes to purification), and the improvement is only mild for the $EB$ case, very significant gains can be obtained for the $BB$ power spectrum, reducing uncertainties by up to a factor of $\sim10$ on large scales.
      \begin{figure*}
        \centering
        \includegraphics[width=0.99\columnwidth]{val_cl_lss_flat}
        \includegraphics[width=0.99\columnwidth]{val_cl_cmb_flat}
        \caption{{\sl Left and Right:} the analogues of the left panels of Figures \ref{fig:lss_val_cls} and \ref{fig:cmb_cls} for the flat-sky LSS and CMB validation sets respectively.} \label{fig:val_cls_flat}
      \end{figure*}

@@

@@

    \nmt~does not currently produce a covariance matrix estimate. This
    is done for several reasons. First, for many problems, such as
    large-scale structure power spectrum estimation, the covariance
    matrix contains import contributions from the connected four-point
    function that depend on the precise nature of the problem and go
    beyond the estimation step that \nmt~strives to do well. Second,
    with all additional options included, the exact analytical
    estimates are either computational infeasible or overly
    approximate. Since majority of the forthcoming experiments will
    have a large number of mock datasets available for testing
    systematic effects, these maps also offer a natural path towards
    estimating the data covariance.

    \nmt~is open-source software and publicly available at \url{https://github.com/LSSTDESC/NaMaster}. To maximize the
    usefulness of this software for the community, it is accompanied
    by an extensive documentation and example code\footnote{See \url{
        https://github.com/LSSTDESC/NaMaster}.}. \nmt~is written in
    {\tt C} and {\tt OpenMP}-parallelized to maximize its performance,
    although we encourage its use through the companion {\tt python}
    wrapper, the fastest-growing language of choice in the analysis of
    astronomical data. The code is in constant development, and we
    will strive to implement further functionality to e.g. estimate
    covariance matrices, extend the range of supported sky
    pixelizations and enhance its computational performance.

\subsection*{Acknowledgements}
We would like to thank Rupert Allison, Jo Dunkley, Franz Elsner, Boris
Leistedt, Thibaut Louis, Mathew Madhavacheril, Sigurd N\ae ss, Hiranya
Peiris, Naomi Robertson and Ben Thorne for useful comments and
discussions. DA acknowledges support from the Beecroft trust and from
the Science and Technology Facilities Council (STFC) through an Ernest
Rutherford Fellowship, grant reference ST/P004474/1. This research
used resources of the National Energy Research Scientific Computing
Center (NERSC), a U.S. Department of Energy Office of Science User
Facility operated under Contract No. DE-AC02-05CH11231.

@@

\section{Linear contamination from non-linear
  contamination} \label{app:lincont}
\as{It is up to David to decide whether he buys this argument or
  not. I'm fine with not and then we soften then claim in 2.2. Doesn't
  really affect any of the paper otherwise.}
Imagine a contaminant field $c$ (e.g. a point-spread function (PSF) size)
that affects the observable field $o$ (e.g. the number count
densities) through a non-linear, but \emph{local} on some scale local
functional $F$:
\begin{equation}
  \delta_o(\nv) \rightarrow \delta_o + F [ \delta_c ]
\end{equation}
where we can, without loss of generality, take out the mean effects
(i.e. there is some mean PSF size and fluctuations around this mean
size produce additional fluctuations in the number density of
fluctuations).

@@

  \section{Flat skies and Fourier transforms}\label{app:flat}
    \subsection{Spin-\texorpdfstring{$s$}{s} fields in flat sky}\label{app:flat.fields}
      In the flat sky we will label a position in the plane by two
      coordinates $(x,y)\equiv{\bf x}$. These coordinates can be
      directly related to increments in the spherical coordinates
      $(\theta,\varphi)$ by considering a sufficiently small map
      centered around the equator (i.e. $\sin\theta\sim 1$). In
      this case we identify the $x$ coordinate with a latitude shift
      ($\delta\theta=-\delta x$) and the $y$ coordinate with longitude
      shifts ($\delta\varphi=\delta y$) \todo{make a plot of
        this?}. \as{no, but isn't it other way round? Latitute is $y$
        in my brain? :)} \dam{it isn't, have a look here: \url{https://healpix.jpl.nasa.gov/html/intronode12.htm}.} Using this dictionary, the differential operator $\eth$ now takes the form:
      \begin{equation}
        \eth= (\partial_x-i\partial_y),\hspace{12pt}\bar{\eth}= (\partial_x+i\partial_y).
      \end{equation}
      The analog of the standard spherical harmonics for flat skies are plane waves $e^{i{\bf k}{\bf x}}$, and the action of $\eth$ on these is:
      \begin{equation}
        \eth^s e^{i{\bf k}{\bf x}}=(ik)^se^{-i\,s\varphi_k}e^{i{\bf k}{\bf x}},\hspace{12pt}
        \bar{\eth}e^{i{\bf k}{\bf x}}=(ik)^se^{i\,s\varphi_k}e^{i{\bf k}{\bf x}}
      \end{equation}

@@

    \subsection{Discrete description and DFTs}\label{app:flat.dft}
      As we discuss in \todo{where}, it is more convenient to describe power spectrum estimation methods in a discretized flat sky. Let the patch of the sky under inspection be contained by a rectangle of sides $L_x$ and $L_y$ (in units of radians), and let us discretize this rectangle by dividing it into an $N_x\times N_y$ grid with pixels of area $\Delta{\bf x}^2\equiv\Delta x\Delta y=(L_x/N_x)(L_y/N_y)$. Each pixel in this grid is then labeled by a pair of integers ${\bf n}\equiv(n_x,n_y)$, and is assigned coordinates ${\bf x}_{\bf n}\equiv(n_x\Delta x,n_y\Delta y)$. Each field in the pixelized map ${\bf a}({\bf x})$ is therefore defined for $n_x\in[0,N_x-1]$, $n_y\in[0,N_y-1]$.

>>>>

      At this stage, the pseudo-$C_\ell$ method proceeds as usual, applying Eq. \ref{eq:pcldef} to the purified field. The purification process however requires a slight modification of the analytical form of the mode-coupling matrix. Specifically, if either $E$ or $B$ modes of a spin-2 field have been purified Eqs. \ref{eq:m02} and \ref{eq:m22} must be modified by exchanging one factor of
      \begin{equation}\nonumber
        \wtj{\ell}{\ell'}{\ell''}{2}{-2}{0}
      \end{equation}
      for
      \begin{align}\nonumber
        &\wtj{\ell}{\ell'}{\ell''}{2}{-2}{0}+2\frac{\beta_{\ell,2}}{\beta_{\ell,1}\beta_{\ell'',1}}\wtj{\ell}{\ell'}{\ell''}{1}{-2}{1}\\
        &\,\,\,+\frac{\beta_{\ell,2}}{\beta_{\ell'',2}}\wtj{\ell}{\ell'}{\ell''}{0}{-2}{2}.
      \end{align}
      per purified field.

@@

      The full covariance matrix is very close to diagonal, as shown in the right panel of Fig. \ref{fig:lss_val_cls}. Under the assumption that the estimator is unbiased ($\langle{\bf s}\rangle={\bf t}$), the values of $\chi^2_i$ should follow a $\chi^2$ distribution with a $N_{\rm dof}$ degrees of freedom. Figure \ref{fig:lss_val_chi2} shows the distribution of $\chi^2_i$ across simulations approximating the covariance matrix as diagonal (blue histograms), compared with the predicted $\chi^2$ distribution for $N_{\rm dof}=681$ (marked by the vertical dashed lines) for the six different power spectrum combinations. Both distributions
      are found to be compatible. Moreover, the dot-dashed lines show the values of $\chi^2$ found for the mean residual with respect to the input power spectra (middle panel of Fig. \ref{fig:lss_val_cls}). The probability to exceed (PTE) associated with these $\chi^2$ values are higher than $10\%$ in all cases, and therefore our validation suite does not show any statistically significant bias in the estimator. This bias, if any, must be at least $\sqrt{1000}\sim30$ times smaller than the statistical uncertainties associated with our validation suite, which are representative of the statistical power of next-generation galaxy surveys.

@@

    \subsection{Curved-sky CMB \texorpdfstring{$B$}{B}-mode validation}\label{ssec:validation.cmb_full}
      \begin{figure*}
        \centering
        \includegraphics[width=0.99\columnwidth]{val_covar_cmb_sph_nocont}
        \includegraphics[width=0.99\columnwidth]{val_covar_cmb_sph}
        \caption{{\sl Left:} covariance matrix for power spectra computed in the absence of foreground contaminants computed from the 1000 simulations of our curved-sky CMB validation suite. The incomplete sky coverage causes a noticeable anti-correlation of neighboring bandpowers at low $\ell$ for the three power spectra ($EE$, $EB$ and $BB$), but beyond that, the covariance matrix is close to diagonal. {\sl Right:} the same as the left panel but this time for simulations containing foreground contaminants. The combination of $B$-mode purification and contaminant deprojection produces noticeable correlations at low $\ell$, especially in the $BB$ spectrum, that must be taken into account.} \label{fig:cmb_val_cov}
      \end{figure*}
      \begin{figure*}
        \centering
        \includegraphics[width=0.80\textwidth]{val_chi2_cmb_sph}
        \caption{Distribution of $\chi^2$ values found for the 1000 simulations in our curved-sky CMB validation suite (blue histograms). Results are shown for the three polarized spectra ($EE$, $EB$ and $BB$), labeled in the top right corner of each panel. The simulated distributions agree well with the expected $\chi^2$ distribution (solid line) assuming Gaussian statistics for $N_{\rm dof}=43$ degrees of freedom (denoted by the dashed vertical lines). The dot-dashed vertical lines mark the $\chi^2$ value found for the mean residual with respect to the input power spectrum (see lower half of the left panel in Fig. \ref{fig:cmb_val_cls}). The associated PTE values are $\gtrsim0.4$, and thus no statistically significant bias is found by our validation suite.} \label{fig:cmb_val_chi2}
      \end{figure*}
      We have carried out a similar set of tests on an ensemble of 1000 simulations from the CMB validation suite described in Section \ref{sssec:validation.suite.cmb}. In our fiducial run we both deprojected the foreground contaminants and subsequently used $B$-mode purification to decouple $E$ and $B$-modes at the map level, since this is the main feature of the code we want to validate. The left panel of Figure \ref{fig:cmb_val_cls} shows the mean of the $EE$, $EB$ and $BB$ power spectra computed for all simulations (solid lines) compared with the expectation from the input spectra for $EE$ and $BB$ (dashed lines, mostly lying under the solid ones). The average residuals normalized over the error on the mean, shown in the lower panel of the same figure, fluctuate around zero with an amplitude of up to $3\sigma$, as would be expected in the absence of a bias in the estimator.

      The above figure was created using $E/B$ mode purification. Disabling this option has no impact on the bias of the estimates, but increases its variance as we show in the Figure \ref{fig:cmb_val_sigma}. We plot diagonal error bars with (red) and without (blue) $B$-mode purification for the $BB$ (solid), $EB$ (dashed) and $EE$ (dot-dashed) power spectra. Although there is no practical improvement in the error bars for the $EE$ spectrum (there is even a slight increase due to the loss of the ambiguous $E$ modes to purification), and the improvement is only mild for the $EB$ case, very significant gains can be obtained for the $BB$ power spectrum, reducing uncertainties by up to a factor of $\sim10$ on large scales.
      \begin{figure*}
        \centering
        \includegraphics[width=0.99\columnwidth]{val_cl_lss_flat}
        \includegraphics[width=0.99\columnwidth]{val_cl_cmb_flat}
        \caption{{\sl Left and Right:} the analogues of the left panels of Figures \ref{fig:lss_val_cls} and \ref{fig:cmb_cls} for the flat-sky LSS and CMB validation sets respectively.} \label{fig:val_cls_flat}
      \end{figure*}

@@

    Besides describing and validating the code, and in order to generalize the pseudo-$C_\ell$ estimator to all the cases supported by \nmt, this paper has also presented a number of results that are, to the best of our knowledge, new. These include:
    \begin{itemize}
      \item Extending the analytical treatment of contaminant deprojection presented in \cite{2016MNRAS.456.2095E} to fields of arbitrary spin.
      \item Combining contaminant deprojection and $E/B$ purification.
      \item Consistently deriving all of this functionality (standard pseudo-$C_\ell$, deprojection and purification for arbitrary spins) in the flat-sky approximation.
    \end{itemize}

@@

    \nmt~does not currently produce a covariance matrix estimate. Although computationally efficent analytical estimates of the Gaussian covariance matrix can be derived under certain approximations \citep{2004MNRAS.349..603E,2005MNRAS.360.1262B}, this is not widely supported by the code for several reasons. First, for many problems, such as large-scale structure power spectrum estimation, the covariance matrix contains important contributions from the connected four-point function that depend on the precise nature of the problem and go beyond the estimation step that \nmt~strives to do well. Second, with all additional options included, especially if a large number of contaminant templates are deprojected, the exact analytical estimates are either computational unfeasible or overly approximate. Since a majority of the forthcoming experiments will rely on a large number of mock datasets available for testing systematic effects, these maps also offer a natural path towards estimating the data covariance.

    \nmt~is open-source software and publicly available at \url{https://github.com/LSSTDESC/NaMaster}. To maximize the usefulness of this software for the community, it is accompanied by an extensive documentation and example code\footnote{See \url{https://github.com/LSSTDESC/NaMaster}.}. \nmt~is written in {\tt C} and {\tt OpenMP}-parallelized to maximize its performance, although we encourage its use through the companion {\tt python} wrapper, the fastest-growing language of choice in the analysis of astronomical data. The code is in constant development, and we will strive to implement further functionality to e.g. estimate covariance matrices, extend the range of supported sky pixelizations and enhance its computational performance.

\subsection*{Acknowledgements}
We would like to thank Rupert Allison, Scott Dodelson, Jo Dunkley, Franz Elsner, Boris Leistedt, Francois Lanusse, Thibaut Louis, Mathew Madhavacheril, Sigurd N\ae ss, Hiranya Peiris, Naomi Robertson and Ben Thorne for useful comments and discussions. DA acknowledges support from the Beecroft trust and from the Science and Technology Facilities Council (STFC) through an Ernest Rutherford Fellowship, grant reference ST/P004474/1. This research used resources of the National Energy Research Scientific Computing Center (NERSC), a U.S. Department of Energy Office of Science User Facility operated under Contract No. DE-AC02-05CH11231.

@@

  \section{Linear contamination from non-linear contamination} \label{app:lincont}
    Consider a contaminant field $c$ (e.g. a point-spread function (PSF) size) that affects the observable field $o$ (e.g. the number count densities) through a non-linear, but \emph{local} on some scale local functional $F$:
    \begin{equation}
      \delta_o(\nv) \rightarrow \delta_o + F [ \delta_c ]
    \end{equation}
    where we can, without loss of generality, take out the mean effects (i.e. there is some mean PSF size and fluctuations around this mean size produce additional fluctuations in the number density of fluctuations).

@@

  \section{Flat skies and Fourier transforms}\label{app:flat}
    \subsection{Spin-\texorpdfstring{$s$}{s} fields in flat sky}\label{app:flat.fields}
      In the flat sky we will label a position in the plane by two coordinates $(x,y)\equiv{\bf x}$. These coordinates can be directly related to increments in the spherical coordinates $(\theta,\varphi)$ by considering a sufficiently small map centered around the equator (i.e. $\sin\theta\sim 1$). In this case we identify the $x$ coordinate with a latitude shift ($\delta\theta=-\delta x$) and the $y$ coordinate with longitude
      shifts ($\delta\varphi=\delta y$) the differential operator $\eth$ now takes the form: %\todo{make a plot of this?}. \as{no, but isn't it other way round? Latitute is $y$ in my brain? :)} \dam{it isn't, have a look here \url{https://healpix.jpl.nasa.gov/html/intronode12.htm}.} Using this dictionary,       
      \begin{equation}
        \eth= (\partial_x-i\partial_y),\hspace{12pt}\bar{\eth}= (\partial_x+i\partial_y).
      \end{equation}
      The analog of the standard spherical harmonics for flat skies are plane waves $e^{i{\bf k}{\bf x}}$, and the action of $\eth$ on these is:
      \begin{equation}
        \eth^s e^{i{\bf k}{\bf x}}=(ik)^se^{-i\,s\varphi_k}e^{i{\bf k}{\bf x}},\hspace{12pt}
        \bar{\eth}e^{i{\bf k}{\bf x}}=(ik)^se^{i\,s\varphi_k}e^{i{\bf k}{\bf x}}
      \end{equation}

@@

    \subsection{Discrete description and DFTs}\label{app:flat.dft}
      As we discuss in \ref{sssec:maths.flat.pcl101}, it is more convenient to describe power spectrum estimation methods in a discretized flat sky. Let the patch of the sky under inspection be contained by a rectangle of sides $L_x$ and $L_y$ (in units of radians), and let us discretize this rectangle by dividing it into an $N_x\times N_y$ grid with pixels of area $\Delta{\bf x}^2\equiv\Delta x\Delta y=(L_x/N_x)(L_y/N_y)$. Each pixel in this grid is then labeled by a pair of integers ${\bf n}\equiv(n_x,n_y)$, and is assigned coordinates ${\bf x}_{\bf n}\equiv(n_x\Delta x,n_y\Delta y)$. Each field in the pixelized map ${\bf a}({\bf x})$ is therefore defined for $n_x\in[0,N_x-1]$, $n_y\in[0,N_y-1]$.
